### 2023-12-27 sections/2-methods.tex
edits in methods

In this manuscript, we are primarily concerned with $k$-NN search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$ of cardinality $|\textbf{X}| = n$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation, for example, the neural-network embedding of an image, the genome of an organism, a radio frequency signal, etc.

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number.
We require that the distance function be symmetric (i.e., $f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$) and has identity of indiscernibles (i.e. $f(x, y) = 0 \iff x = y$).
If, in addition to these constraints, the distance function obeys the triangle inequality (i.e. $f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then the distance function is also a \emph{distance metric}.
Under distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics while Cosine distance is not a metric because it does not obey the triangle inequality.

The choice of distance function varies by dataset and domain.
For example, with neural-network embeddings, one could use Euclidean (L2) or Cosine distance.
With genomic or proteomic data, Levenshtein or Hamming distances are more appropriate.
With time-series data, one could use Dynamic Time Warping (DTW) distance.

@@

where $B_X(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset.
To illustrate the difference, consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575.
Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws of physics that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional manifold within the 8,575-dimensional embedding space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets obey the manifold hypothesis.

\subsection{Clustering}
\label{subsec:methods:clustering}

@@

The \emph{metric entropy} ($\mathcal{N}_{r}(X)$) for a given cluster radius $r$ was defined, in~\cite{yu2015entropy}, as the minimum number of clusters of a uniform radius $r$ needed to cover all data.
In this paper, we define the metric entropy $\mathcal{N}_{\hat{r}}(X)$ of a dataset $X$ for the hierarchical clustering scheme as the number of leaf clusters in the tree where $\hat{r}$ is the mean radius of all leaf clusters.

We start by performing a divisive hierarchical clustering on the dataset using CLAM.
The procedure is similar to that outlined in CHESS, but with the following improvements:
better selection of poles for partitioning (see Algorithm~\ref{alg:methods:clustering:partition}) and depth-first reordering of the dataset (see Section~\ref{subsubsec:methods:clustering:dataset-depth-first-reordering}).

CAKES assumes the manifold hypothesis. 
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$. 
While we sometimes use Euclidean notions, such as voids, volumes and proximity, to talk about the geometric and topological properties of clusters and of the manifold, CLAM and CAKES do not rely on such notions; 
they serve merely as a convenient and intuitive way to discuss the underlying mathematics.

\subsubsection {Cluster Partitioning}
\label{subsubsec:methods:cluster-partitioning}

@@

The \emph{radius} of $C$ is the maximum distance from the \emph{center} to any other point in $C$.

The point which is responsible for that radius (i.e., the furthest point from the \emph{center}) is designated the \emph{left pole}.
The point which is furthest from \emph{left pole} is designated the \emph{right pole}.

We then partition the cluster into a \emph{left child} and a \emph{right child}, where the left child contains all points in the cluster which are closer to the \emph{left pole} than to the \emph{right pole}, and the right child contains all points in the cluster which are closer to the \emph{right pole} than to the \emph{left pole}.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.

Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria, for example, a minimum cluster radius, a minimum cluster cardinality, a maximum tree depth, etc.
This process is described in Algorithm \ref{alg:methods:clustering:partition}.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Partition(\emph{C, criteria})} % give the algorithm a caption
    \label{alg:methods:clustering:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $C$, a cluster
        \REQUIRE $criteria$, user-specified stopping criteria

        \STATE $seeds \Leftarrow \lfloor \sqrt{|C|} \rfloor$ random points from $C$
        \STATE $c \Leftarrow$ geometric median of $seeds$
        \STATE $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \STATE $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \STATE $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \STATE $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$
        \IF{$|L| > 1$ \textbf{and} $L$ satisfies $criteria$}
            \STATE Partition($L$)
        \ENDIF
        \IF{$|R| > 1$ \textbf{and} $R$ satisfies $criteria$}
            \STATE Partition($R$)
        \ENDIF
    \end{algorithmic}
\end{algorithm}

\subsubsection {Depth-First Reordering}
\label{subsubsec:methods:clustering:dataset-depth-first-reordering}

CAKES also improves upon CHESS by reordering the dataset in depth-first order of traversal of the tree.

In CHESS, each cluster stored a list of indices into the dataset.
This list was used to retrieve the clusters' points during search.
Though this approach allowed us to retrieve the points in constant time, its memory cost was prohibitively high.
With a dataset of cardinality $n$ and each cluster storing a list of indices for its points, we stored a total of $n$ indices at each depth in the tree.
Assuming a balanced tree, and thus $\mathcal{O}(\log(n))$ depth, this approach had a memory overhead of $\mathcal{O}(n\log(n))$.

@@

In this work, we introduce a new approach wherein, after building the cluster tree, we reorder the dataset so that points are stored in a depth-first order.
Then, within each cluster, we need only store its \emph{cardinality} and an \emph{offset} to access the its points from the dataset.

The root cluster has an offset of zero and a cardinality equal to the number of points in the dataset.
A left child has the same offset as that of its parent, and the corresponding right child has an offset equal to the left child's offset plus the left child's cardinality.

@@

% \subsubsection {Scaling Behavior of Cluster Radii}
% \label{subsubsec:methods:clustering:guaranteed-decrease-in-cluster-radii}

% While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:methods:clustering:partition}), unfortunately, this assumption is incorrect. 
% Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii;
% in particular, we prove in this section that cluster radii will stop increasing after at most $d$ partitions, where $d$ is the fractal dimensionality of the dataset.

@@

% In this case, Partition will select a maximally distant pair of points to use as poles;
% without loss of generality, say it chooses two points at the extrema one of the $d$ axes. 
% After one partition, the maximum pairwise distance along that axis will be halved, i.e. bounded above by $R$.
% The next recursive Partition will select another of the $d$ axes, and once again, the distance along that axis will be bounded above by $R$
% Thus, after at most $d$ partitions, the maximum pairwise distance along each axis will be bounded above by $R$. 
% The overall (i.e., not restricted to one axis) maximum pairwise distance 
% will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes. 
% See Figure~\ref{fig:methods:scaling-behavior} for an example.

@@

% \begin{figure}[ht!]
%     \centering
%     \includegraphics[width=3.4in]{images/geometry/geometry.pdf}
%     \caption{
%         Scaling behavior of radii on a uniform-density disk, which represents the worst-case scenario for a two-dimensional distribution. 
%         $R_0$ denotes the radius of the root cluster $C_0$, and $o_0$ denotes its center. 
%         After one application of Algorithm~\ref{alg:methods:clustering:partition}, we have $C_1$, with radius $R_1$ and center $o_1$.
%         The right triangle formed by $o_0$, $o_1$, and $y_+$ in $C_1$ shows that $R_0 < R_1$.
%         Hence, the radius of a child cluster can be larger than its parent.
%         However, after another application of Algorithm~\ref{alg:methods:clustering:partition}, we have consumed all $d$ orthogonal axes, 
%         as shown in $C_2$.
%         Now, clearly $R_2 < R_0$.
%         In fact, for each subsequent application of Algorithm~\ref{alg:methods:clustering:partition}, the radius of the resulting cluster is bounded above by $\frac{\sqrt{2}R}{2}$.
%     }
%     \label{fig:methods:scaling-behavior}
% \end{figure}

\subsubsection {Complexity}
\label{subsubsec:methods:clustering:clustering:complexity}

@@

\subsection{The Search Problem}
\label{subsec:methods:the-search-problem}

Given Algorithm~\ref{alg:methods:clustering:partition}, we can now pose the $k$-NN and $\rho$-NN search problems.

Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-NN search aims to find the set $S_q$ such that $|S_q| = k$ and $\forall x \in \textbf{X} \setminus S_q$, $\ f(x, q) > \max\{f(y, q): y \in S_q \}$;
in other words, $k$-NN finds the $k$ closest points to $q$ in $\textbf{X}$.
We also have the $\rho$-NN search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are no more than a distance $\rho$ from $q$.

Given a Cluster $C$, let $c$ be its center and $r$ be its radius. Our $\rho$- and $k$-NN algorithms make use of the following properties, as illustrated in Figure~\ref{fig:methods:deltas}.

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.pdf}
    \caption{$\delta$, $\delta^{+}$, and $\delta^{-}$ for a cluster $C$ and a query $q$.}
    \label{fig:methods:deltas}
\end{figure}

@@

\subsection{\texorpdfstring{$\rho$}{p}-Nearest Neighbors Search}
\label{subsec:methods:rnn-search}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/overlapping-children-3.pdf}
    \caption{
        $m$ denotes the midpoint of $\overline{l r}$. $q'$ is the projection of $q$ onto $\overline{l r}$.
        $d$ is the distance from $q'$ to $m$.
        If $d \leq \rho$, both child clusters could contain points in the query ball.
        Otherwise, only the right child cluster could contain such points. 
    }
    \label{fig:methods:overlapping-children-3}
\end{figure}

To determine whether both children can contain points in the query ball, we consider Figure~\ref{fig:methods:overlapping-children-3}.
Here, we overload the notation for $\overline{x y}$ to refer to the line segment joining points $x$ and $y$ as well as the length of that line segment.

Let $q$ denote the query, $\rho$ denote the search radius, and $l$ and $r$ denote the cluster's left and right poles respectively (see Section~\ref{subsubsec:methods:cluster-partitioning}).
Without loss of generality, we assume that $\overline{q r \vphantom{l}} \leq \overline{q l}$ (i.e., the query is closer to $r$ than to $l$).

Now let $q'$ be the projection of $q$ to $\overline{l r}$, $m$ be the midpoint of $\overline{l r}$ and $d$ be the distance from $q'$ to $m$.
From Algorithm~\ref{alg:methods:clustering:partition}, we assign a point in the parent cluster to the left child,
Thus, if if $q'$ is too far from $m$, i.e. $\rho < d$, then when the left child cannot contain points inside the query ball, we proceed only with the right child.
Otherwise, we proceed with both children.

@@

\begin{equation}
    d = \overline{m q' \vphantom{l}} = \overline{m r \vphantom{l}} - \overline{q' r \vphantom{l}} = \frac{\overline{l r}}{2} - \overline{q' r \vphantom{l}}.
    \label{eq:methods:d-expression}
\end{equation}

Let $\theta$ denote $\angle l r q$, as shown in Figure~\ref{fig:methods:overlapping-children-3}.
By the Law of Cosines on $\triangle l r q$, we have that

\begin{equation}
    \text{cos}(\theta) = \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r} \cdot \overline{q r \vphantom{l}}}.
    \label{eq:methods:cosine-law}
\end{equation}

@@

\begin{equation}
    \text{cos}(\theta) = \tfrac{\overline{q' r \vphantom{l}}}{\overline{q r \vphantom{l}}}.
    \label{eq:methods:cosine-law-right-triangle}
\end{equation}

Combining Equations~\ref{eq:methods:cosine-law} and~\ref{eq:methods:cosine-law-right-triangle}, we have that

\begin{equation}
    \overline{q' r \vphantom{l}} = \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}}.
    \label{eq:methods:q-prime-r}
\end{equation}

Substituting Equation~\ref{eq:methods:q-prime-r} into Equation~\ref{eq:methods:d-expression}, we have that

@@

\begin{equation}
    d \leq \rho \iff (\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho.
    \label{eq:methods:check-d}
\end{equation}

Note, in particular, that Equation~\ref{eq:methods:check-d} only requires distances between real points, and so it can be used with any distance function, even when $q'$ and $m$ are not real points or cannot be imputed from the data.

To perform $\rho$-nearest neighbors search, we first perform a coarse tree-search, as outlined in Algorithm~\ref{alg:methods:rnn:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained search, as outlined in Algorithm~\ref{alg:methods:rnn:leaf-search}, to find all points in which are within a distance $\rho$ of the query.

\begin{algorithm} 
    \caption{tree-search($C$, $q$, $\rho$)} 
    \label{alg:methods:rnn:tree-search} 
    \begin{algorithmic}
        \REQUIRE $C$, a cluster
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

        \STATE $c$ $\Leftarrow$ \emph{center} of $C$
        \STATE $r$ $\Leftarrow$ \emph{radius} of $C$

        \IF{$C$ is a leaf \textbf{or} $f(q, c) + r \leq \rho$}
            \STATE \textbf{return} $\{C\}$
        \ELSIF{Inequality~\ref{eq:methods:check-d} is true}
            \STATE $L$ $\Leftarrow$ \emph{left child} of $C$
            \STATE $R$ $\Leftarrow$ \emph{right child} of $C$
            \STATE \textbf{return} tree-search($L, q, \rho$) $\cup$ tree-search($R, q, \rho$)
        \ELSIF{$f(q, l) < f(q, r)$}
            \STATE \textbf{return} tree-search($L, q, \rho$)
        \ELSE
            \STATE \textbf{return} tree-search($R, q, \rho$)
        \ENDIF
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{leaf-search($S$, $q$, $\rho$)} 
    \label{alg:methods:rnn:leaf-search} 
    \begin{algorithmic}
        \REQUIRE $S$, a set of clusters
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

@@

        \FOR{$C \in S$}
            \STATE $c$ $\Leftarrow$ \emph{center} of $C$
            \STATE $r$ $\Leftarrow$ \emph{radius} of $C$
            \IF{$f(q, c) + r \leq \rho$}
                \STATE $H$ $\Leftarrow$ $H \cup \{C\}$
            \ELSE
                \FOR{$p \in C$}
                    \IF{$f(p, q) \leq \rho$}
                        \STATE $H$ $\Leftarrow$ $H \cup \{p\}$
                    \ENDIF
                \ENDFOR
            \ENDIF
        \ENDFOR

@@

\begin{algorithm} 
    \caption{$\rho$-NN-search($root$, $q$, $\rho$)} 
    \label{alg:methods:rnn:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

@@

The asymptotic complexity of $\rho$-nearest neighbors is the same as in~\cite{ishaq2019clustered} and shown in Equation~\ref{eq:methods:rnn:complexity}.

\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{\log_2 \mathcal{N}_{\hat{r}}(X)}_{\textrm{metric entropy}} +
        \overbrace{ \left| B_X(q, \rho) \right|}^{\textrm{output size}}
        \underbrace{ \left( 1 + \frac{2 \cdot \hat{r}}{ \rho} \right) ^ d}_{\textrm{scaling factor}}
    \Bigg)
    \label{eq:methods:rnn:complexity}
\end{gather}

>>>>

In this manuscript, we are primarily concerned with $k$-NN search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$ of cardinality $|\textbf{X}| = n$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation, for example, the neural-network embedding of an image, the genome of an organism, a measurement of a radio frequency signal, etc.

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a non-negative real number.
A distance value of zero defines an identity among points (i.e. $f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among the points.
We also require that the distance function be symmetric (i.e., $f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e. $f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \emph{distance metric}.
Under distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics while Cosine distance is not a metric because it does not obey the triangle inequality (e.g. consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on a Cartesian plane).

The choice of distance function varies by dataset and domain.
For example, with neural-network embeddings, one could use Euclidean (L2) or Cosine distance.
With genomic or proteomic data, Levenshtein, Smith-Waterman, Needleman-Wunsch or Hamming distances are useful.
With time-series data, one could use Dynamic Time Warping (DTW) or Wasserstein distance.

@@

where $B_X(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset.
To illustrate the difference, consider the Silva 18S rRNA dataset which contains which contains genomes with unaligned lengths of up to 3,718 base pairs.
Hence, the \emph{embedding dimension} of this dataset is at least 3,718.
However, due to physical constraints (namely, the laws that govern RNA replication and mutation), the data are constrained to a lower-dimensional manifold within this embedding space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets obey the manifold hypothesis.

\subsection{Clustering}
\label{sec:methods:clustering}

@@

The \emph{metric entropy} ($\mathcal{N}_{r}(X)$) for some radius $r$ was defined, in~\cite{yu2015entropy}, as the minimum number of clusters of a uniform radius $r$ needed to cover the data.
In this paper, we define the metric entropy $\mathcal{N}_{\hat{r}}(X)$ of a dataset $X$ for the hierarchical clustering scheme as the number of leaf clusters in the tree where $\hat{r}$ is the mean radius of all leaf clusters.

We start by performing a divisive hierarchical clustering on the dataset using CLAM.
The procedure is similar to that outlined in CHESS, but with the following improvements:
better selection of poles for partitioning (see Algorithm~\ref{alg:methods:partition}) and depth-first reordering of the dataset (see Section~\ref{sec:methods:clustering:depth-first-reordering}).

CAKES assumes the manifold hypothesis.
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$.
While we sometimes use Euclidean notions, such as voids, volumes and proximity, to talk about the geometric and topological properties of the clusters and manifold, CLAM and CAKES do not rely on such notions;
they serve merely as convenient and intuitive vocabulary to discuss the underlying mathematics.

\subsubsection{Building the Tree}
\label{sec:methods:clustering:building-the-tree}

CLAM starts by building a binary tree of clusters using a divisive hierarchical clustering algorithm.

@@

The \emph{radius} of $C$ is the maximum distance from the \emph{center} to any other point in $C$.
The point which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole} and the point which is furthest from left pole is designated the \emph{right pole}.

We then partition the cluster into a \emph{left child} and a \emph{right child}, where the left child contains all points in the cluster which are closer to the left pole than to the right pole, and the right child contains all points in the cluster which are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.

Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria, for example, minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc.
This process is described in Algorithm \ref{alg:methods:partition}.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Partition($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $C$, a cluster
        \REQUIRE $criteria$, user-specified stopping criteria

        \STATE $seeds \Leftarrow \lfloor \sqrt{|C|} \rfloor$ random points from $C$
        \STATE $c \Leftarrow$ geometric median of $seeds$
        \STATE $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \STATE $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \STATE $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \STATE $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$

        \IF{$|L| > 1$ \textbf{and} $L$ satisfies $criteria$}
            \STATE Partition($L$, $criteria$)
        \ENDIF

        \IF{$|R| > 1$ \textbf{and} $R$ satisfies $criteria$}
            \STATE Partition($R$, $criteria$)
        \ENDIF
    \end{algorithmic}
\end{algorithm}

Note that Algorithm~\ref{alg:methods:partition} does not necessarily produce a balanced tree.
In fact, for real datasets, we would expect anything but a balanced tree.
The imbalance in the tree would be due to the varying density of the sampling in different regions of the manifold and the low dimensional ``shape'' of the manifold itself.
The only case in which we would expect a balanced tree is if the dataset were uniformly distributed, e.g. in a $d$-dimensional hyper-cube.
In this sense, the imbalance in the tree is a feature, not a bug, as it reflects the underlying structure of the data.

\subsubsection{Depth-First Reordering}
\label{sec:methods:clustering:depth-first-reordering}

After building the tree, we reorder the dataset so that the points are stored in the order that they would be visited during a depth-first traversal of the tree.

In CHESS, each cluster stored a list of indices into the dataset.
This list was used to retrieve the clusters' points during search.
Although this approach allowed us to retrieve the points in constant time, its memory cost was prohibitively high.
With a dataset of cardinality $n$ and each cluster storing a list of indices for its points, we stored a total of $n$ indices at each depth in the tree.
Assuming a balanced tree, and thus $\mathcal{O}(\log(n))$ depth, this approach had a memory overhead of $\mathcal{O}(n\log(n))$.

@@

In this work, we introduce a new approach wherein, after building the cluster tree, we reorder the dataset so that points are stored in a depth-first order.
Then, within each cluster, we need only store its \emph{cardinality} and an \emph{offset} to access the its points from the dataset.
The root cluster has an offset of zero and a cardinality equal to the number of points in the dataset.
A left child has the same offset as that of its parent, and the corresponding right child has an offset equal to the left child's offset plus the left child's cardinality.

@@

% \subsubsection{Scaling Behavior of Cluster Radii}
% \label{sec:methods:clustering:scaling-behabior-of-cluster-radii}

% While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:methods:partition}), unfortunately, this assumption is incorrect. 
% Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii;
% in particular, we prove in this section that cluster radii will stop increasing after at most $d$ partitions, where $d$ is the fractal dimensionality of the dataset.

@@

% In this case, Partition will select a maximally distant pair of points to use as poles;
% without loss of generality, say it chooses two points at the extrema one of the $d$ axes. 
% After one partition, the maximum pairwise distance along that axis will be halved, i.e. bounded above by $R$.
% The next recursive Partition will select another of the $d$ axes, and once again, the distance along that axis will be bounded above by $R$
% Thus, after at most $d$ partitions, the maximum pairwise distance along each axis will be bounded above by $R$. 
% The overall (i.e., not restricted to one axis) maximum pairwise distance 
% will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes. 
% See Figure~\ref{fig:methods:radii-scaling-behavior} for an example.

@@

% \begin{figure}[ht!]
%     \centering
%     \includegraphics[width=3.4in]{images/geometry/geometry.pdf}
%     \caption{
%         Scaling behavior of radii on a uniform-density disk, which represents the worst-case scenario for a two-dimensional distribution. 
%         $R_0$ denotes the radius of the root cluster $C_0$, and $o_0$ denotes its center. 
%         After one application of Algorithm~\ref{alg:methods:clustering:partition}, we have $C_1$, with radius $R_1$ and center $o_1$.
%         The right triangle formed by $o_0$, $o_1$, and $y_+$ in $C_1$ shows that $R_0 < R_1$.
%         Hence, the radius of a child cluster can be larger than its parent.
%         However, after another application of Algorithm~\ref{alg:methods:clustering:partition}, we have consumed all $d$ orthogonal axes, 
%         as shown in $C_2$.
%         Now, clearly $R_2 < R_0$.
%         In fact, for each subsequent application of Algorithm~\ref{alg:methods:clustering:partition}, the radius of the resulting cluster is bounded above by $\frac{\sqrt{2}R}{2}$.
%     }
%     \label{fig:methods:radii-scaling-behavior}
% \end{figure}

\subsubsection{Complexity}
\label{sec:methods:clustering:complexity}

@@

\subsection{The Search Problem}
\label{sec:methods:the-search-problem}

Given Algorithm~\ref{alg:methods:partition}, we can now pose the $k$-NN and $\rho$-NN search problems.

Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $\forall x \in \textbf{X} \setminus H, \ f(x, q) > \max \{ f(y, q): y \in H \}$;
in other words, $k$-NN finds the $k$ closest points to $q$ in $\textbf{X}$.
We also have the $\rho$-NN search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are no more than a distance $\rho$ from $q$.

Given a Cluster $C$, let $c$ be its center and $r$ be its radius.
Our $\rho$-NN and $k$-NN search algorithms make use of the following properties, as illustrated in Figure~\ref{fig:methods:deltas}.

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.8]{images/geometry/deltas.pdf}
    \caption{{\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.}
    \label{fig:methods:deltas}
\end{figure}

@@

\subsection{\texorpdfstring{$\rho$}{p}-Nearest Neighbors Search}
\label{sec:methods:rnn-search}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.8]{images/geometry/overlapping-children-3.pdf}
    \caption{The geometry of a query ball overlapping with a cluster and either one or both of its children.}
    \label{fig:methods:overlapping-children}
\end{figure}

To determine whether both children can contain points in the query ball, we consider Figure~\ref{fig:methods:overlapping-children}.
Here, we overload the notation for $\overline{x y}$ to refer to the line segment joining points $x$ and $y$ as well as the length of that line segment.

Let $q$ denote the query, $\rho$ denote the search radius, and $l$ and $r$ denote the cluster's left and right poles respectively (see Section~\ref{sec:methods:clustering:building-the-tree}).
Without loss of generality, we assume that $\overline{q r \vphantom{l}} \leq \overline{q l}$.
Now let $q'$ be the projection of $q$ to $\overline{l r}$, $m$ be the midpoint of $\overline{l r}$ and $d$ be the distance from $q'$ to $m$.
As a consequence of how we assign a point in the parent cluster to the left child in Algorithm~\ref{alg:methods:partition}, if $q'$ is too far from $m$, i.e. $\rho < d$, then when the left child cannot contain points inside the query ball.
In such a case we proceed only with the right child.
Otherwise, we proceed with both children.

@@

\begin{equation*}
    d = \overline{m q' \vphantom{l}} = \overline{m r \vphantom{l}} - \overline{q' r \vphantom{l}} = \frac{\overline{l r}}{2} - \overline{q' r \vphantom{l}}.
\end{equation*}

Let $\theta$ denote $\angle l r q$, as shown in Figure~\ref{fig:methods:overlapping-children}.
By the Law of Cosines on $\triangle l r q$, we have that

\begin{equation*}
    \text{cos}(\theta) = \tfrac{\overline{l r}^2 + \ \overline{q r \vphantom{l}}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r} \cdot \overline{q r \vphantom{l}}}.
\end{equation*}

@@

\begin{equation*}
    \text{cos}(\theta) = \tfrac{\overline{q' r \vphantom{l}}}{\overline{q r \vphantom{l}}}.
\end{equation*}

Combining the previous two equations and solving for $\overline{q' r \vphantom{l}}$, we have that

\begin{equation*}
    \overline{q' r \vphantom{l}} = \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}}.
\end{equation*}

Substituting for $\overline{q' r \vphantom{l}}$ in the equation for $d$, we have that

@@

\begin{equation}
    d \leq \rho \iff (\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho.
    \label{eq:methods:overlapping-children}
\end{equation}

Note, in particular, that Equation~\ref{eq:methods:overlapping-children} only requires distances between real points, and so it can be used with any distance function, even when $q'$ and $m$ are not real points or cannot be imputed from the data.

To perform $\rho$-NN search, we first perform a coarse \emph{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained \emph{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points in which are no more than a distance $\rho$ from the query.

\begin{algorithm} 
    \caption{tree-search($C$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search:tree-search} 
    \begin{algorithmic}
        \REQUIRE $C$, a cluster
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

        \IF{$C$ is a leaf \textbf{or} $\delta^+_C \leq \rho$}
            \STATE \textbf{return} $\{C\}$
        \ELSE

            \STATE $[l, r]$ $\Leftarrow$ \emph{pole}s of $C$
            \STATE $[L, R]$ $\Leftarrow$ \emph{children} of $C$

            \IF{$\overline{q l} < \overline{q r \vphantom{l}}$}
                \STATE $[r, l] \Leftarrow [l, r]$
                \STATE $[R, L] \Leftarrow [L, R]$
            \ENDIF

            \IF{$(\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho$}
                \STATE \textbf{return} tree-search($L, q, \rho$) $\cup$ tree-search($R, q, \rho$)
            \ELSE
                \STATE \textbf{return} tree-search($R, q, \rho$)
            \ENDIF
        \ENDIF
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{leaf-search($S$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search:leaf-search} 
    \begin{algorithmic}
        \REQUIRE $S$, a set of clusters
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

@@

        \FOR{$C \in S$}
            \IF{$\delta^+_C \leq \rho$}
                \STATE $H$ $\Leftarrow$ $H \cup \{C\}$
            \ELSE
                \FOR{$p \in C$}
                    \IF{$f(p, q) \leq \rho$}
                        \STATE $H$ $\Leftarrow$ $H \cup \{p\}$
                    \ENDIF
                \ENDFOR
            \ENDIF
        \ENDFOR

@@

\begin{algorithm} 
    \caption{$\rho$-NN-search($root$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

@@

The asymptotic complexity of $\rho$-NN is the same as in~\cite{ishaq2019clustered} and shown in Equation~\ref{eq:methods:rnn-search-complexity}.

\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{\log_2 \mathcal{N}_{\hat{r}}(X)}_{\textrm{metric entropy}} +
        \overbrace{ \left| B_X(q, \rho) \right|}^{\textrm{output size}}
        \underbrace{ \left( 1 + \frac{2 \cdot \hat{r}}{ \rho} \right) ^ d}_{\textrm{scaling factor}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}
